\label{section:introduction}

Security protocols are used every day by billions of users and applications to guarantee a certain degree of security and privacy over communications happening on the (insecure) Internet (SSH \cite{rfc4251} and TLSv1.3 \cite{TLSv1.3_specs} protocols are just a few famous examples). Although, there is a catch: designing such protocols has been proven to be very error-prone. As an example, consider the Needham-Schroeder public-key protocol \cite{NSPK}, which has been believed to be secure for almost 20 years - before a fatal flaw was found and corrected \cite{NSPK_LoweGavin}.

Given the importance of the correctness of such protocols and the difficulties for designers to ensure it, it has been necessary to \textit{formally} prove the absence of security vulnerabilities. Towards this aim, a set of tools has been developed to assist the designing of a new protocol.

In this thesis, we are going to compare two tools: Tamarin prover and Proverif. In particular, we will compare them for the formal verification of Telegram's protocol: MTProto2.0. The protocol has already been analyzed and formalized with Proverif by \MMNV{} \cite{MTProto2-Proverif-impl}, and it has been discussed in the associated paper \cite{MTProto2-Proverif}. The formalization in Tamarin will be based on this work, with remarkable differences due to the diverse nature of the two tools.

In chapter 2, we will present the symbolic and computational models. The two tools, Tamarin prover and Proverif, will be briefly described in chapter 3. In the next chapter, we will also present the model of a very simple protocol with both tools. A description of the MTProto2.0 protocol will be given in chapter 5, while, in chapter 6, we will describe its formalization in Tamarin prover. Finally, in chapter 7, we will compare Tamarin prover and Proverif.